Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-03-14 output/paper/manuscript.qmd
Re-writing discussion basically.

===
To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.

@@

González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@wechsler1949wechsler] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.

@@

BPA

Few studies have looked into different classes of non-persistent {{< acr edcs >}}. Shoaff et al. investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites (MEHP, MEHHP, MEOHP, MECPP, MnBP, MiBP, MBzP, monohydroxyisobutyl phthalate (MHiBP), monocarboxyoctyl phthalate (MCOP), monoisononyl phthalate (MNP), and monohydroxybutyl phthalate (MHBP)), especially in boys [@ShoaffCoullWeuve:2020].

Oh et al.
[@OhKimKannan:2023]

Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment (too broad) in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].
===

>>>>

To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.

@@

González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@wechsler1949wechsler] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.

@@

Among phenols, some studies provide preliminary evidence of an association between BPA and {{< acr adhd >}} in children aged 8 to 15 years [@TewarAuingerBraun:2016] and in a case-control study of children aged 6 to 12 years [@LiZhangKuang:2018], especially in boys. Except for working memory, there does not seem to be evidence of an association between BPA and cognitive abilities in Spanish boys aged 9 to 11 years [@Rodriguez-CarrilloMustielesPerez-Lobato:2019].
Few studies have looked into different classes of non-persistent {{< acr edcs >}}. Shoaff et al., for instance, investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites (MEHP, MEHHP, MEOHP, MECPP, MnBP, MiBP, MBzP, monohydroxyisobutyl phthalate (MHiBP), monocarboxyoctyl phthalate (MCOP), monoisononyl phthalate (MNP), and monohydroxybutyl phthalate (MHBP)), especially in boys [@ShoaffCoullWeuve:2020].

TBD

Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment (too broad) in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].

TBD
